An anthropological account of the Vim editor:\\ ouroboros, features and tweaks after 10 years of usage

### 2017-12-04 article/article.tex
updating vimrc and article

\documentclass{article}
\usepackage{graphicx}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{hyperref}
\hypersetup{
	%pagebackref=true,
	pdfcreator={LaTeX with abnTeX2},
	pdfkeywords={abnt}{latex}{abntex}{USPSC}{trabalho acadêmico}, 
	colorlinks=true,       		% false: boxed links; true: colored links
	linkcolor=blue,          	% color of internal links
	citecolor=blue,        		% color of links to bibliography
	filecolor=magenta,      		% color of file links
	urlcolor=blue,
	allbordercolors=black,
	bookmarksdepth=4
}
\usepackage[utf8]{inputenc}
\usepackage{tipa}
\newcommand{\ttt}[1] {
	\texttt{<#1>}}
\newcommand{\tttt}[1]{\texttt{#1}}
% \usepackage{fontspec}
% \usepackage[mathletters]{ucs}
% \newcommand{\ttt}[1] {
% 	\texttt{<#1>}}
% \newcommand{\tttt}[1]{\texttt{#1}}

\begin{document}
\title{Basic concepts and tools for the Toki Pona minimalist language:
Wordnet synsets; analysis, synthesis and syntax highlighting of texts}
% Anthropological computation
\author{Renato Fabbri\\
\texttt{renato.fabbri@gmail.com}\\
University of São Paulo,\\
Institute of Mathematical and Computer Sciences\\
São Carlos, SP, Brazil
}
\maketitle
\begin{abstract}
  A minimalist constructed language (conlang)
  is useful for experiments and is comfortable for making tools.
  The Toki Pona conlang is minimalist both in the vocabulary
  (only 14 letters, X phonemes, and 120 words + 4 synonyms)
  and in the $\approx10$ syntax rules.
  The language is useful for being a used and somewhat established
  minimalist conlang with at least hundreds of speakers.
  In this article, we describe current concepts and resources
  for Toki Pona,
  and make available Python scripted routines for
  analysis and synthesis of texts and the specification of syntax highlighting schemes.
  We focus on the analysis of the basic vocabulary,
  as corpus analyses were found in~\cite{corpus}.
  The synthesis is based on the basic sentence template,
  relates to context by keeping track of used words,
  and renders larger texts by using a fixed number of phonemes (e.g. for poems)
  and number of sentences, words and letters (e.g. for paragraphs).
  Syntax highlighting is performed in 24bit true colors, currently only
  with the Vim text editor.
  Word colors reflect morphosyntactic classes given in the official dictionary
  and different solutions to necessary and arbitrary choices are described and implemented.
  In summary, this text holds potentially novel conceptualizations about,
  and tools and results in analyzing, synthesizing or syntax highlighting the
  Toki Pona language.
  % subj (li), obj (e), prepositions, qualification (word concatenation), 
  % sent (la), particles, mi-sina, pi, punctuation is freestyle, pre-verb,
\end{abstract}
{\bf keywords:} Natural Language Processing, Textual synthesis, Syntax highlighting, Wordnet, Toki Pona

\section{Introduction}\label{intro}
Toki Pona is a minimalist conlang (constructed language)
with only 124 words (120 without the synonyms).
Therefore, the concepts are usually very general and different,
and, without context, the words are rarely related through
meronymy and hyponymy.
Such a linguistic setting is desired because of the simplicity
which entails easier e.g. learning and tool making.
Another reason why the minimalistic language design is compelling
is the study and harnessing of the strong and weak forms of
the Sapir-Whorf hypothesis (linguistic relativity),
i.e. that language influences or dictates ones thought
and world experience.
Accordingly, one uses a conlang as a thinking tool (or platform)
or to make experiments about the influence language has on
the thoughts of the `speaker' (also writer and reader).
Toki Pona is often described as a tool to meditate,
to simplify the thinking processes, and as a way to modify
the mood and impressions about the world.~\cite{tp1,tp2,tp3,tp4}
In~\cite{interview}, Sonja Lang (the creator of Toki Pona),
describes that she has seen the language been used successfuly
in the context of management, creation of texts, legal texts,
etc.

In this article, we present a conceptual overview of the language,
which is fit both to the newcomer and to the expert.
We also summarize
texts and software gadgets
for Toki Pona that
were developed by us and by other Toki Pona users.
Most importantly, the conceptual overview is considerably different
from what we found in the literature~\cite{pu,pije,memrise},
with emphasis on simplicity and flexibility,
and the software routines yield analyses, syntheses
and highlighting Toki Pona what was thought to be complementary
to the software currently available.

Next subsections hold a description of the general resources available,
a historical note, and some words about natural and constructed languages.
Section~\ref{basic} describes Toki Pona's phonology and syntax.
Section~\ref{hacks} presents the software routines we made available
and immediate results, such as listings and statistics of words,
poems and short stories, and coloring schemes.
Conclusions and further work are in Section~\ref{conc}.
Appendix~\ref{mytoki} holds considerations about my usage of Toki Pona,
with thoughts of rule breaking and potentially new conlangs.
Appendix~\ref{ftp} holds final words in Toki Pona.

% definition, historic, status (groups, speakers, etc).

\subsection{Resources on Toki Pona}
One might organize current resources for the Toki Pona
language in: references and learning material, corpus,
websites, interaction groups (here users talk a post texts and comments), and software gadgets.
The main references of the language are:
  the official book ``Toki Pona: The Language of Good'',
    authored by Sonja Lang, the creator of the language;
the online book ``o kama sona e toki pona!'',
from jan Pije~\cite{kama}.
For a more comprehensive view of the resources available
for the user, we suggest following the links 
from~\cite{wikiToki}.
Section~\cite{third} holds a summary of what we found in software
gadgets related to Toki Pona.

\subsection{Historical note}
Toki Pona was developed as an internal and personal language
by Sonja Lang~\cite{interview}.
It was released as a draft in 2001 and in 2007 some documents
reported it to have a few hundred speakers.
The English official book~\cite{lipu} was released only in 2014.
In 2016, a version of the official book was released in French.
Nowadays, one finds a number of texts about Toki Pona and written
in it, and uses of Toki Pona for artificial intelligence~\cite{roila}
and software tools~\cite{tptoll1,tptool2,tptool3},
in social platforms such as Facebook groups~\cite{fbtp,fbtpt},
microblogging~\cite{tokilili}, Telegram~\cite{telegram} and IRC~\cite{irc}.

\subsection{Natural and constructed and artificial languages}
A language one uses (or might use) to communicate by speaking
and writing is called `natural language'.
A `constructed language' is a natural language built by someone or
a group, such as Esperanto, Toki Pona, and Lojban.
An artificial language is a language yield by artificial agents,
such as in AI routines, and might be considered
within `cultural evolution' studies.
Formal languages are defined by tokens and rules to
operate them, they span from computer programming languages
to math and formal models for natural languages.
Lojban is a constructed language with the purpose of being
an artificial language because it is more axiomatic.

Constructed languages are in some traditions called artificial
languages, but creators most often prefer to use the term `standardized'
or 'planned' language with the argument that the conlang
is rooted on natural languages, and 'artificial' is misleading~\cite{conlanWikip}.
The preferred terms seem to be planned or constructed languages
or conlang.
The construction of languages is called glossopoeia.
Toki Pona is a conlang which might be classified
as engineered for experiments, meditation and philosophy;
and suitable for auxiliary international language and
as an artistic language. 
% taxonomy, vocabulary, ontology, terminology,

% conlangs, natural langs, programming langs, Lojban as an
% intermediary
% non-verbal languages, musical corporal language, 
% non-verbal languages confront Sapir-Whorf hypothesis.

\section{Overview of the language}\label{basics}
This section describes very succinctly the formation of
words and sentences in the Toki Pona language.
It should enable a newcomer to grasp the essentials
of the language and the experienced to
acquire new insights.
Furthermore, it is a solid reference of the phonological and syntax
rules.

\subsection{Phonology}
Words in Toki Pona are written using only 14 letters:
\begin{itemize}
  \item Vowels a (open), e (mid front), o (mid back), i (close front),
    u (close back).
  \item Consonants j, k, l, m, n, p, s, t, w:
    \begin{itemize}
      \item Nasal: m (labial), n (coronal).
      \item Plosive: p (labial), t (coronal), k (dorsal).
      \item Fricative: s (coronal).
      \item Approximant: w (labial), l (coronal), j (dorsal).
    \end{itemize}
\end{itemize}

There are standard guidelines for pronunciation,
but the language allows for considerable allophonic
variation.
For example, /p t k s l/ might be pronounced
[p t k s l] or [b d g z \textipa{\!R}].
Especially for poetry, one might consider
j and w to be vowels
(e.g. j as 'i' and w as 'u').

Syllables are of the form (C)V(N):
an optional consonant, a vowel and an optional nasal consonant.
Non word-initial syllables must follow the pattern CV(N).
The following sequences are forbidden: ji, wu, wo, ti, nm, nn.

\subsection{Syntax}
\subsubsection{Fundamental notions}
As in other natural languages, 
colloquial Toki Pona might have
incomplete sentences and deviate from the
norm.
The basic structure of sentences is:
`subject' (Noun) li `predicate' (Verb) e `object' (Noun).
The li might be repeated to associate more than
one predicate to the subject.
The particle li is omitted if the subject is a simple mi (I or us)
or sina (you). A discussion about issues (problems)
yield by this rule
and how I deal with them is in Appendix~\ref{mytoki}.

The e might be repeated to associate more than
one object to a predicate.
Sentences might be related though la,
'sentence' la 'sentence', where the second sentence is
the main sentence, and the first sentence is a condition
to the first.
Multiple la-s are not described in literature,
but I assume that one might assume the last sentence
being a conditional to the next,
except in cases where the context strongly suggests
otherwise.

Noun and verb phrases are (usually) built with the non-particle words.
The first word is the noun or verb and subsequent words
qualify the noun or verb.
The pi particle might be used to separate sequences of words
to be evaluated before the relation yield by pi.
As pi is often ill understood and used,
the following structures might be handy for newbies and as a
reference:
\begin{itemize}
  \item No pi, `word word word':
word $\leftarrow$ (qualifies 1) word $\leftarrow$ (qualifies 2)  word.
  \item One pi, 'word pi word word': word $\leftarrow$ (qualifies 2) [
      word $\leftarrow$ (qualifies 1)  word ].
  \item Two pi-s: `word pi word word word pi word word':
    word $\leftarrow$5 [word 2 word] 3 word $\leftarrow$4 word 1  word;
or:
    word $\leftarrow$5 [word 1 word] 2 word $\leftarrow$4 word 3  word.
\end{itemize}

Notes on the usage of pi:
\begin{itemize}
  \item In a sequence of words, without pi, the second word qualifies
    the first, the third word qualifies the phrase yield by the first
    two words, the fourth word qualifies the noun yield by the first
    three words and so on.
  \item It is redundant to use pi before the last word in a noun or
    verb phrase if there is no other pi, reason why it is most often
    omitted.
    Its use in this case is regarded as wrong~\cite{lipu,pije},
    but, as one might notice, it does not add (much) information
    through syntax (order of qualifies is conserved).
    It adds as an emphasis because of greater length of the written segment, as a preparation: 'jan lili pi mama' (mommy's child).
    Also used in~\cite{akesiWawa}.
  \item The book by jan Pije~\cite{kama} describes another use for pi:
    after li to mean possession, e.g. `soweli li pi sina' (your pet).
    This employment of pi might be regarded as correct, but are promptly written
    as a noun phrase (e.g. `soweli sina') and is not mentioned
    by the official book~\cite{tpLabg}.
\end{itemize}

All the words except the structural particles (li, e, la, pi)
are usable in any position in noun and verb phrases,
(i.e. 116 words).
Notice that the phrase expresses a noun in a noun phrase (subject or
object) or a verb (in the predicate).
And that the first word of the phrase is the noun or verb,
and that subsequent words are adjectives or adverbs.
The pi and other particles restarts the phrase to
first word being a noun.

At this point, the only missing syntax rule is related to
the prepositions: kepeken, lon, sama, tan, tawa.
They might appear at the end of phrases,
should be followed by another phrase,
and require no particle. E.g.
`toki *tan* jan Pije li pana e sona *tawa* mi'.
Because prepositions can be used as adjectives
or nouns or verbs, adverbs, etc, and what follows
a preposition complement might also be of any morphosyntactic class,
they are core sources of ambiguity.
three:
* preposition
* absence of li on sina and mi
* Incomplete sentences
(ona li) tawa lon kon e ilo pi suli mute

\subsubsection{Particles}
Beyond the structural particles (li, e, la, pi) presented
in last section, other particles are:

\begin{itemize}
  \item a or kin, emphasis.
  \item o, vocative or imperative ('jan lukin sitelen o, li wawa')
  \item taso, means however as sentence or 'only' if adjective.
  \item anu, en: 'or' and 'and'. Used for nouns in nous phrases.
    For repetition of verbs, repeat li.
    For object nouns, repeat e.
    If the noun is complementing a preposition (tawa, lon),
    one might repeat the preposition.
    As Toki Pona is a recent language, and is able to cope with
    variation due to its simplicity, I would advocate for
    using en and anu wherever there is no ambiguity.
    E.g.: 'mi tawa anu moku lon tenpo lili?'
    In the official documentation that is not described and
    might be regarded as wrong in strict canonical.
  \item nanpa, denotes numbering.
  \item seme, for questions, used next to the thing being asked for.
    'Why?' might be expressed as
    'seme la sina pana e moku lon sewi'.
  \item mu, for animal noises. For me it is not a particle, as in the
    official dictionary, but a noun.
    I also like to use it as a verb:
    'mi pakala e luka. mu mute.'
\end{itemize}

The vocabulary specifies morphosyntactic classes:
nouns, adjectives, verbs, pre-verbs, adverbs, particles, prepositions, and numbers.
I find that they might help the user and newcomer, but
it might also suggest a deviation from what I understand and read:
the words might be used indistinctly to be the nouns
(subjects, the predicate when there is no object, objects,
and preposition complements),
the adjectives (anything that does not start the noun phrase or follows a pi),
verbs
(follows mi or sina or li or a preposition),
adverbs (follows the verb).
The pre-verbs (wile, ken, awen, kama, lukin, sona),
might follow a verb, but might also be understood
as the verb qualified with the next word,
which carries a very similar if not identical meaning.
The pre-verbs are all also defined as other morphosyntactic classes,
such as adjective, noun, verb.
The only exception is wile, which is only a pre-verb.

Thus, the classes given in the dictionary dictate little
in practice:
jan kala li lape lon ni.
Where kala, lape and ni are in this phrase
as adjective, verb and noun,
and are in the dictionary as noun,
adjective and adjective.

As far as I can see, one should regard
the particles li, e, pi, and la
and punctuation.
The other tokens of the vocabulary
might be used in any of the remaining positions.

\subsubsection{Convention for recognizing POS tags by speaker and machine}

\begin{itemize}
  \item Noun: the first word in a noun phrase.
    After an 'e' and after a pi,
    The first word in the sentence if
    sentence does not start by the verb.
    Might be in the position of the verb
    if the sentence has no object.
  \item Adjective: second word on after an e and after
    Second word on in the subject phrase if present.
  \item Verb: after a li, mi or sina.
    If there is no object, the verb position
    is often a noun.
  \item After a preposition,
    there can be a noun phrase, a verb phrase
    or nothing.
  \item Notice that there is ambiguity in the structure
    introduced by the omission of li after mi and sina.
    Also, when there is no object, a noun or a verb
    or an adjective might be in the verb position
    if there is no object.
    The prepositional complement is also not defined.
    (mi moku tawa pali, tawa tomo).
    So, these are sources of syntactic ambiguities
    in Toki Pona.
    They might be solved or minimized by using the semantics
    of the words.
    One preliminary effort in this direction might be
    using the classes in the official dictionary to resolve
    ambiguities whenever possible.
    This solution is not optimal in correct POS tagging,
    and does not solve all possible ambiguities
    (there are words classified as nouns and adjectives,
    adjectives and verbs, nouns and verbs, particle and verb).

    Another source of ambiguity is the pre-verbs as described
    in the literature~\cite{tpLang,pije}.
    But I find it reasonable to understand them as verbs.

>>>>

\documentclass{article}
\usepackage{graphicx}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{hyperref}
\hypersetup{
	%pagebackref=true,
	pdfcreator={LaTeX with abnTeX2},
	pdfkeywords={abnt}{latex}{abntex}{USPSC}{trabalho acadêmico}, 
	colorlinks=true,       		% false: boxed links; true: colored links
	linkcolor=blue,          	% color of internal links
	citecolor=blue,        		% color of links to bibliography
	filecolor=magenta,      		% color of file links
	urlcolor=blue,
	allbordercolors=black,
	bookmarksdepth=4
}
\usepackage[utf8]{inputenc}
\newcommand{\ttt}[1] {
	\texttt{<#1>}}
\newcommand{\tttt}[1]{\texttt{#1}}

\begin{document}
\title{An anthropological account of the Vim editor:\\
ouroboros, features and tweaks after 10 years of usage}
% Anthropological computation
\author{Renato Fabbri\\
\texttt{renato.fabbri@gmail.com}\\
University of São Paulo,\\
Institute of Mathematical and Computer Sciences\\
São Carlos, SP, Brazil
}
\maketitle
\begin{abstract}
The Vim editor is very rich in capabilities
and thus complex.
This article is a description of the Vim text editor
and a set of considerations on its usage and design.
It is the result of more than ten years of experience
with Vim for writing and editing various types of documents,
e.g.:
Python, C++, JavaScript, ChucK, programs;
and \LaTeX, Markdown, HTML, RDF, Make and other markup files;
binary files.
It is said that it takes about ten years to master (or start mastering)
this text editor, and I find that other experienced users
have a different view of Vim and that they use a different
set of features.
Therefore, this document exposes my insights in order
to confront my usage with that of other Vim users.
Another goal is
to make available a reference document with which new users
can grasp a sound overview by reading it and the discussions that
it might generate.
Also, it should be useful for users of any degree of experience,
including me, as a compendium of commands, namespaces and tweaks.
Upon feedback, and maturing of my Vim usage,
this document might be enhanced or receive additional
material.
\end{abstract}

\section{Introduction}\label{intro}
Vim is a very complex editor,
considered by the Linux community to be
matched only by Emacs.
They both are the standard advanced text editors
of the free software and open source communities
and have been developed for decades.
This document describes the Vim editor
and proposes a set of enhancements of the user
experience through simple tweaks.
The editor has a very mature documentation
and the contents herein presented is a
report on the overall understandings I
have of Vim after a little bit more
of ten years using it.
The purposes of this document are:
\begin{itemize}
  \item to help new users in grasping Vim essentials
  and convenient practices.
  \item To attain a sound overall description of the editor.
  \item To record the view of the editor that
  a user (me) has after $10$ years of usage.
  \item To confront my usage with that of other experienced
  users. This is helpful for me, but also for the other users
  as they might benefit from this content and from discussions
  that might arise from it.
  \item To propose some enhancements to Vim through simple tweaks and potential plugins.
\end{itemize}

Advanced users might just skim throughj
Section~\ref{basics} and consider more carefully
Section~\ref{namespaces}.
The concluding remarks and proposed enhancements
might also deserve some attention.

I've had experience with other editors, e.g. Kate, gedit, and Notepad2.
I used Vim for writing and editing computer code (Python, Javascript, C++, ChucK, bash, etc), markup languages (HTML, CSS, RDF, Markdown, \LaTeX, etc) and binary files.
Eventually, I edited database files and other types os files.
With Vim, I mostly write software (web and scientific),
music, poems and short stories.
It is very useful because:
\begin{itemize}
  \item it is meant to be a plain text (e.g. ascii, utf8) editor
  and does not (by standard) insert special charaters (e.g. for formating, with binary instructions).
  \item It has a powerful architecture and set of commands.
  \item It is highly configurable and most often the users
  have a set of commands for standard settings and hacks kept in the vimrc and other configuration files.
\end{itemize}

The standard capabilities of the editor
should become clear in Section~\ref{basics}.
Vim is constantly evolving and there are many plugins,
some of them very popular for both general and specific
users.
Accordingly, there are many possible enhancements,
and Section~\ref{conc} report the most prominent of them
for me and potential workarounds made available as plugins.

This document is written is a DRY KISS
(Don't Repeat Yourself, Keep It Simple Stupid) style.
Complex is to master the use of Vim
and one finds sound
references in help files and a nice vimrc.
Therefore the following content should be kept
as uncomplicated and original as possible.
Also, because of Vim's complexity and entailed
bond of this document to my usage,
there is an anthropological component
which is evident in the use of the first person
in sentences.
This can be understood as anthropological computation
or physics~\cite{anPh,anPh2} and observed to help
in the technological groundwork of the civil society.

% Vi and Vim
\subsection{Historical note}
Vim was first released publicly in 1991.
It is a cross-platform GNU licensed free and open source extended clone of Bill Joy's vi text editor.
Vim's development is coordinated (and performed) since the beginning
mainly by Bram Moolenaar.
Today, current bleeding-edge version is 8.0.1257.
I found no explicit stable, alpha or beta versions.
I found no scientific article on Vim 
(this might be the first one), although there are
books, software and third-party documentation on the web.

\section{Basics}\label{basics}
Vim's interface is text-based.
In the GUI mode (gVim),
there are convenient menus and toolbars
but all functionalities are still available though
the command line mode.
Vimscript is the internal language of Vim,
and is often used for scripting by users
although other languages might be used 
(e.g. Python, Perl, Lua, Racker, Ruby and Tcl). 
Each line of a Vimscript is a command on the
command-line mode.
This section can be thought as a tutorial
that focuses on the namespaces, i.e. sets of tokens
that carry values or triggers procedures.
One should see Appendix~\ref{keys}
to understand the notation vim uses.

\subsection{The bare minimum}\label{minimum}
You open a file with Vim by executing
the command: \texttt{vim <filename>}.
Inside Vim, you start in the normal
mode, and might want to move around using
\texttt{h-j-k-l} for left-down-up-right.
To insert characters, move your
cursor to the desired location an press i,
which puts Vim in the insert mode.
Go back to normal mode by pressing
\texttt{<ESC>} or \texttt{<C-C>}.
You save the file by typing \texttt{:w<CR>},
and exit Vim by typing \texttt{:q<CR>}.
You can save and quit with \tttt{:wq<CR>}
or \tttt{:x<CR>} or \tttt{ZZ}.

\subsection{Vim help}
You can find help for vim in various places.
The standard resource is the Vim help files.
They are accessed by typing \texttt{:h <anything><CR>}
in normal mode.
Examples of such \texttt{<anything>} are:
color, navigation, :vs, vimtutor.
Type 
\texttt{:h usr\_toc<CR>}
to access the official User Manual,
which is considerably lengthy and complex
and is usually not read by users for a few years.
In learning Vim, one
might want to run the \texttt{vimtutor} command
(outside Vim) to start the Vim Tutor.

There are good resources on the Web for learning
and tweaking Vim:
\begin{itemize}
  \item ``Vim Adventures'' is an online RPG game for practicing
  and memorizing Vim commands while having fun.
  This game is quite famous among Vim users.
  \item There are official and semi-official Vim sites e.g.:
  \url{www.vim.org}, \url{https://www.vi-improved.org} and
  \url{http://vim.wikia.com/}.
  \item Many hacks, understandings and general issues
  (e.g. how to make such a move) are asked and answered
  in online platforms (e.g. Quora, Stack Overflow, Stack Exchange, Reddit, Email list, IRC Channel).
  One often finds these links through a search engine.
  \item You can find many videos about Vim.
  One traditional site is \url{http://vimcasts.org},
  but you might find them by a search engine (e.g. \url{http://derekwyatt.org/vim/tutorials/}) or in Youtube and Vimeo.
\end{itemize}

% any other commands? files?
\subsection{Namespaces}\label{namespaces}
Vim is a text editor ouroboros~\cite{ouroWiki} because
text and writing alters text and writing.
You have namespaces where tokens have scalar
or complex values.

\subsubsection{Commands and mappings}
There are commands, typing sequences which trigger automated actions,
for each mode:
\begin{itemize}
	\item in Normal mode all keys are mapped to commands.
		There is redundancy and additional commands
		using Ctrl and Shift keys.
		Some keys expect a second key,
    and have combinations not used (available for new mappings),
		specially the z and g.
    See Section~\ref{normal} and Appendix~\ref{notes}
    for more insights into the commands available in the
    normal mode.
	\item In the other modes, the sequences available for mappings are more obvious and abundant.
		One should look at \tttt{:h index} to know about all the standard mappings
		and use \tttt{:map} to list the user-defined mappings.
\end{itemize}

\ttt{C-\textbackslash} is often reserved for extensions,
which makes it a safe namespace to use (while there are no
such extensions).

A colon command can be written as a string
and executed by the \tttt{:execute} colon command.
E.g. \tttt{:execute 'vs afile.txt'}.
As there are colon commands that execute commands in other
modes, e.g. \tttt{:normal ?\^def },
the \tttt{:execute} is a way to build commands in any mode,
e.g.
\tttt{:execute 'normal' tabn 'gt'}.

\subsubsection{Variables}
There are some types of variables in Vim:
\begin{itemize}
	\item Environment variables: names start with \$ and hold system
		variables, such as \tttt{\$PATH} and \tttt{\$PWD}.
	\item Option variables: names start with \& and are meant to control the behavior of the editor.
		One might change a value through set or let, e.g.
		\tttt{:set bg=light} or \tttt{:let \&bg=light}.
	\item Registers: start with @ and are meant for automation and transfer of texts (copy and paste).
	\item Internal variables are created with let and preferably have a prefix:
	\tttt{b:}, \tttt{w:}, \tttt{t:}, \tttt{l:}, \tttt{s:},
		are local to the buffer, window, tab page, function, and
		sourced Vim file, respectively.
 \tttt{v:}, \tttt{g:} are global, the first are predefined by Vim.
		\tttt{a:} is for function arguments.
		If there is no prefix, the variable is global or internal to a function if occurring inside a function.
		More about internal variables in \tttt{:h internal-variables}.
	\item The value of a variable can be a scalar, string, list, dictionary, function reference, etc (see \tttt{:h eval}).
\end{itemize}
You can echo any of such variables or use in expressions.
Notice that you will only be able to echo a \tttt{b:} variable inside
the buffer where it is defined.
For all the Vimscript capabilities, including loops, conditionals,
and builtin functions, refer to \tttt{:h vim-script} and Section~\ref{script}.
Classes are possible only in rudimentary forms, e.g. through dictionaries,
but the language is otherwise overall quite powerful,
specially in dealing with text and editor behavior, as expected.

\subsubsection{State lists}\label{state}
Vim keeps a number of useful lists which expresses the state of the editor:
\begin{itemize}
	\item The entered commands are accessed through \tttt{:hist a}. 
    The tokens a / s : can be used for specific types of commands, such
		as search and colon commands.
	\item File buffers are kept with numeric ids. See buffers with \tttt{:ls} and load a buffer to the window with \tttt{:b <num or token in file name>}.
  \item The windows open are listed in \tttt{:ls} with a character \tttt{a} in the second column, and are listed with \tttt{:tabs}.
	\item Tabs list can be reached through \tttt{:tabs}.
		It is usual both to show and hide the tabs bar (mapping in~\cite{vimrc}).
	\item Jumps are available through \tttt{:jumps}.
		One positions the cursor at each jump through \ttt{C-o} and \ttt{C-i}.
  \item Registers are available through \tttt{:reg} command, as variables and through shortcuts in different modes.
    They also keep track or your copy, edition and deletion and are promptly defined
    by recording a typing sequence with the \tttt{q} normal command.
    Vim keeps only the last edition, in register \tttt{".}.
    An autocommand to keep the four latest inserts is in~\cite{vimrc}.
    A hack to keep the latest deletions and copies in the standard register
    might follow the same pattern, but use another \tttt{:h event} and
    monitor register \tttt{""} (maybe also \tttt{"0}).
	\item An undo list can be accessed with \tttt{:changes}.
	\item A list with all the sourced scripts in a Vim instance is displayed through \tttt{:scriptnames}.
	\item The markers defined are listed with the \tttt{:marks} command.
		These are set by \tttt{mX} in normal mode, where X is the marker identifier.
		Uppercase letters are cross buffers.
	\item Quickfix and Location lists which are populated through \tttt{:vim} and \tttt{:make}
		and variations, such as \tttt{:grep}.
		One might run \tttt{:vim /section/ \%} and then \tttt{:copen}
		to open the Quickfix window, where the lines of occurrence are in sequence
		and one can \ttt{CR} one of them to have the cursor in the main window active at
		the first character of the match.
		One might run \tttt{:lvim /section/ \%} and then \tttt{:lopen}
		to use the location-list window instead of the Quickfix, which
		is very similar, but one per window instead of one per buffer.
		More information in \tttt{:h quickfix}.
	\item A tags list have to be made so one can use tags.
		Most often one will generate the tags list using the exuberant ctags,
		which supports dozens of languages.
		E.g. \tttt{:!ctags-exuberant functions.py} or \tttt{!ctags-exuberant -R ./},
		and then using \ttt{C-]} to
		go to the position of tag under cursor, and \tttt{:tabe tags} to open
		the tags file.
	\item The argument list holds a list of files to be edited or browsed. 
		The list can be input at Vim startup (e.g. \tttt{\$ vim file1.txt file2.py})
	or using commands (e.g. \tttt{:ar ./*}).
		The file being edited is changed by \tttt{:n} and \tttt{:p} commands,
		one might perform actions on each file in argument list using \tttt{:argdo}.
		All files in argument list are also in the buffers list.
    For further information, see \tttt{:h arglist} and Section~\ref{netrw}.
  \item All the autocommands might be accessed through \tttt{:au}.
    They are event-triggered actions Vim performs.
\end{itemize}

A file with information about the state of the editor
can be achieved through: \tttt{:source \$VIMRUNTIME/bugreport.vim},
and this script might be consulted because it has
a collection of commands to access various settings of Vim.
Another good list of commands to know about Vim's state is kept on
\url{http://vim.wikia.com/wiki/Displaying_the_current_Vim_environment}.
I would specially highlight the \tttt{:syntax} command because
it displays the tokens and related type of meanings when run
inside e.g. a \tttt{.py}, \tttt{.vim} or help file.

\subsubsection{On the persistence of visual cues about the editor state}\label{visual1}

You can keep track of the editor state though commands and persistent visual
cues, specially the tabs bar, the status line, and the line
reserved for the command-line.
For state persistence, one might keep an undo file for each file as
in~\cite{vimrc}.
Sessions are easy to manage, enabling one to save and load the
editor's state, with the opened windows, tabs, buffers, etc.
I use the mappings in~\cite{vimrc} because they keep the sessions
in a reasonable directory and makes it easier to remember and tweak
the standard commands to deal with sessions.
More information in \tttt{:h sessions}.
One might use \tttt{:h views} to keep the state of one window,
but sessions keep all the states from all windows.
This entails a strategy to deal with Vim that is similar to
the use of Byobu/Tmux/Screen.
The main limitation I found to this approach is that
Vim is not keeping track of the terminals opened.
If you open a terminal inside Vim with the \tttt{:term}
command, you will save the session as usual, but when loading
you get dummy empty windows for them and an error message.
Screen and Tmux are the most popular terminal multiplexers.
Byobu, built on top of them, has awesome keyboard shortcuts
for managing sessions, windows and splits of terminals.
Byoby/Screen/Tmux keep the state for future load.

Because browsing the interface in Vim is fast,
and it favors copy and paste of text,
I tend to keeep a tab with some terminals:
one with an IPython shell, another two for compiling latex
and opening PDF files (e.g. with \tttt{\$ evince <filename.pdf>}).
One enters normal mode in terminal with \tttt{C-W N}
and it is very corfortable to copy and browse the bash history of the
terminal.
I found the mappings on~\cite{vimrc} very helpful for directing
editor focus to splits (\ttt{C-hjkl}) and tabs (\tttt{gr}, \tttt{gt}),
which I make available across terminal-job and normal modes.
But I've been thinking on using \ttt{C-} commands also
for tabs (not only for splits).

A good strategy I find is to have selective visual cues of the state to make persistent or hide
and a mapping to toggle each of them.
Now I toggle byobu/screen/tmux bar with \ttt{F5},
status line and tabs bar with \ttt{localleader-T or B} according to script~\cite{vimrc}.
I am mostly using the cleanest setting, toggling on the tabs bar and status line sometimes. Numbering is always there. I rarely turn them off but keep the mappings \ttt{leader-n or N} to toggle just in case.
Instead of keeping the status bar, I use \ttt{C-g} to know about the
file and \ttt{gC-g} to know more and rarily.
It seems not possible to remove the statusbar between horizontal splits.
After asking around and experimenting, I realized that it seems
reasonable to keep at least one line dividing the windows, so
if it comes to it, I just \tttt{set statusline=-}.
Unfortunatelly, as far as I could dig, one will need to enter
Vim code to enable a horizontal split without losing a line.
My ideal of this feature would be to have a visual cue of the
first and/or last line of the windows in the lines-number
column, or complete the spaces and empty chars with \$\$\$\$ or so.

Autocommands are the standard way to define event-trigered routines in Vim.
These are often related to particular file types, but are also often
in defining the automated behavior of Vim.
In~\cite{vimrc} is found an autocommand for keeping track of the last inserted texts
in the \tttt{".lkjh} registers (\tttt{@.lkjh} variables).

\subsection{Using Vim's modes}\label{modes}
Vim has some basic and fully implemented modes of usage:
\begin{itemize}
  \item Normal mode: used for changing
  the position of the cursor or the text displayed
  at the window.
  A core goal of the normal mode it to allow fast
  navigation of the document while allowing
  the typist to maintain its fingers on the home row
  (i.e. on the center of the keyboard).
  The mode is also used for manupulating text
  (e.g. copy, paste, delete, change case) and
    changing to other modes.
  \item Insert mode: for inserting text.
  \item Command-line mode: for entering Ex commands.
  \item Ex mode: similar to command-line mode,
  but more specialized for running various Ex commands.
  \item Visual mode: for making, manipulating and navigating
  selections of texts.
  \item Select mode: similar to visual mode but
  favors CUA\footnote{IBM Common User Access: \url{https://en.wikipedia.org/wiki/IBM_Common_User_Access}.}.
\end{itemize}

There is another basic mode, but it is not fully implemented:
the Terminal-Job mode.
There are seven additional modes which are mostly subordinate 
to the basic modes and will be described when convenient.
The manual page for Vim modes can be accessed by typing
\texttt{:h vim-mode}.
Some of the modes are now further considered for
an overview of the Vim usage possibilites.

\subsubsection{Normal mode}\label{normal}
Sometimes also called navigation or command mode,
the normal mode is most powerful for
navigating, manipulating texts and changing to other modes.

The simplest of these three is changing to other modes:
type any of these letters to change to insert mode:
\texttt{iIaAoOsScC}. More on the transition between
normal and insert mode on Section~\ref{navIn}.
Type any of these characters to change to command-line mode:
\texttt{:/?}.
Type \texttt{Q} to enter Ex mode.
Type \texttt{v}, \texttt{V}, \texttt{CTRL+V} to enter visual mode.
